\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
standardized to bit-rate

\subsection*{Abstract}
In this paper we compare the performance SampleRate and Minstrel, two popular bit rate selection algorithms that have widespread real-world usage on the Linux operating system. We use a novel trace-based approach that allows us avoid in-depth kernel programming and write simple Python scripts instead. We also introduce improvements to the Minstrel algorithm that allow for significant gains in throughput. 
\section{Introduction}

One of the key ways that wireless networks differ from wired is that wireless networks have varying link rates. Link conditions vary with time due to interference from other devices, changing network geometry, and mobile clients. An optimal rate at some time $t$ may be different from the optimal rate just 30 seconds earlier. A good bit rate selection algorithm has to detect and adapt to these conditions. If the chosen bit rate is too slow, then the throughput will be unnecessarily low. If the rate is too high, then failures will be very frequent and throughput will again suffer.

There are three main classes of bit rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols. Frame-based protocols measure the fraction of successfully received packets. SNR protocols make decisions based on the estimated Signal to Noise ratio. Cross-layer protocols use SoftPHY data from the physical layer. The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstraction.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.

@@

In Figure~\ref{figure:2} we show the performance of constant bit rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

Bit rate selection algorithms strive to have a throughput near or greater than the optimal constant rate. For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate. In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.

@@

Bit rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bitrates. SampleRate, however, spends too much time sampling low bit rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

@@

\begin{figure}[htb]
  \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
\label{figure:2}
\end{figure}

@@

Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks. 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities. There is an 802.11n version of Minstrel, but there is no implementation of SampleRate.

>>>>

\subsection*{Abstract}
In this paper we compare the performance SampleRate and Minstrel, two popular bit-rate selection algorithms that have widespread real-world usage on the Linux operating system. We use a novel trace-based approach that allows us avoid in-depth kernel programming and write simple Python scripts instead. We also introduce improvements to the Minstrel algorithm that allow for significant gains in throughput. 
\section{Introduction}

One of the key ways that wireless networks differ from wired is that wireless networks have varying link rates. Link conditions vary with time due to interference from other devices, changing network geometry, and mobile clients. An optimal rate at some time $t$ may be different from the optimal rate just 30 seconds earlier. A good bit-rate selection algorithm has to detect and adapt to these conditions. If the chosen bit-rate is too slow, then the throughput will be unnecessarily low. If the rate is too high, then failures will be very frequent and throughput will again suffer.

There are three main classes of bit-rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols. Frame-based protocols measure the fraction of successfully received packets. SNR protocols make decisions based on the estimated Signal to Noise ratio. Cross-layer protocols use SoftPHY data from the physical layer. The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks as they violate the network layering abstraction.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit-rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.

@@

In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate. For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate. In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.

@@

bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bit-rates. SampleRate, however, spends too much time sampling low bit-rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

@@

\begin{figure}[htb]
  \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
  \caption{Throughput of using a constant 802.11b/g bit-rate in clear line-of-sight from the access point.}
\label{figure:2}
\end{figure}

@@

Very little analysis has been done of the performance of bit-rate selection algorithms on 802.11n networks. 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities. There is an 802.11n version of Minstrel, but there is no implementation of SampleRate.
